Extended Data

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
work on discussion
- delete unneeded notes
- add notes on points to hit

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. r tree_rings`).
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016]. <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.--> <!--Krista, Also note that maxrate_DOY was advanced in RP+DP at SCBI but only DP at HF. Maxrate_DOY for RP at HF was actually significantly delayed. This could be a consequence of the severe drought in 1999 which we've included in our analysis. My brain is completely fried to the point that I barely remember what the literature claims this max rate is related to (water stress, right?). So Maybe you're in a better position to add some discussion about this here. -->
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
*However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.* <!--issue #122-->

@@

- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]

~~~

One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
It is theoretically possible that extra C is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Extra C could also be saved within trees as non-structural carbohydrates (NSC) and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014]. 
Extension of our tree-ring analysis revealed weak correlation between April $T_{max}$ and growth the following year (sig. pos. correlations for 5/142 RP and 3/66 DP species-site combinations, Fig. `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.

>>>>

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. r tree_rings`).
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
*However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.* <!--issue #122-->

@@

- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]
- our study provides further evidence of sink limitation to growth
- we do not know where the extra C is going. If we look at instances where growth is more strongly limited than GPP, we see that NSCs can build up[@refs], and this build up can feed back to reduce photosynthesis[@refs]. We can also look to studies documenting enhanced GPP with no enhancement of growth. In a mature eucalypt forest exposed to elevated CO2, enhanced GPP did not result in higher growth, but rather was released back to the atmosphere on relatively short time scales through enchanced respiration[@jiang_fate_2020]
~~~

One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
It is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Extra C could also be saved within trees as non-structural carbohydrates (NSC) and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014]. 
Extension of our tree-ring analysis revealed weak correlation between April $T_{max}$ and growth the following year (sig. pos. correlations for 5/142 RP and 3/66 DP species-site combinations, Fig. `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.
